---
title: "hydrostreamer tutorial"
author: "Marko Kallio"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{hydrostreamer tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The development of 'hydrostreamer' was inspired by a study, where one component was estimating water availability in rural villages. In that study, water availability was estimated using a regional distributed hydrological model with 5km grid resolution. The problem in the study was how it represented the available water quantity at a village. First, the grid size was too coarse to represent the stream network accurately. A regional model with 5km grid size is quite fine, but for a village level estimates where distance to water source was an important attribute. Field representation (which raster in essence is) was not satisfactory.

Hence, *hydrostreamer* was conceived. *hydrostreamer* takes an off-the-shelf runoff product of an arbitrary resolution, and assigns runoff to an explicitly represented river network. To keep things simple and approachable, there is no runoff modelling involved in the use of *hydrostreamer*, rather, it post-processes results from either hydrological or land surface models and with minimal input requirements. The concept of *hydrostreamer* is very simple and easy to grasp, with a design philosophy that it should be easy to use for anyone, without background in hydrology.

In practise, similar ideas have been used in a number of existing solutions; for instance river routing software HydroROUT, [RAPID](https://github.com/c-h-david/rapid/), or [mizuRoute](https://github.com/NCAR/mizuRoute). These also take off-the-shelf runoff products, and apply river routing algorithms to predict streamflow at explicit (vector) river segments. What sets *hydrostreamer* apart from these solutions is that it is written in R, is easy to install and use, and most importantly, focus is not in river routing, but in the downscaling step.

This tutorial aims to explain the workflow in *hydrostreamer* and showcase some of it's capabilities. More specific tutorials will be added in time, which concentrate on different parts of hydrostreamer

# Using **hydrostreamer**

Some example data have been included to the package for tutorial purposes. It includes 

* 1 by 1 degree DEM at located in Southeast Asia. The DEM is originally ALOS World 3D at 30 meter resolution (Tanado et al 2014), which has been resampled to 0.005 degree resolution. 

* 1 by 1 degree runoff timeseries in the same area as the DEM. Runoff is sourced from the Linear Optimal Runoff Aggregate (LORA) at 0.5 degree resolution (see Hobeichi et al 2019). The unit of runoff is mm/s (kg/m2/s), and are provided with a monthly timestep.

* A river network derived from the provided DEM with 216 river segments.

* River segment specific catchment areas, delineated from the provided DEM. Catchments are provided for a subset (n = 41) of the river segments.

Let's first load the data and inspect it:

```{r message=FALSE, warning=FALSE}
library(hydrostreamer)
library(raster)
library(lubridate)
library(dplyr)
library(ggplot2)
library(tidyr)

data(example_rivers)
data(example_basins)
runoff <- brick(system.file("extdata", "runoff.tif", package = "hydrostreamer"))
data(example_dem)

plot(runoff[[1]]) 
plot(st_union(basins), add=TRUE)
plot(river, add=TRUE)

plot(dem)
plot(st_union(basins), add=TRUE)
plot(river, add=TRUE)

```


*hydrostreamer* main workflow occurs in four steps (and four functions):

1. Convert the runoff timeseries to a HSgrid object
2. Compute weights for each river segment
3. Downscale runoff
4. Apply river routing

## 1. Converting raster runoff to a polygon network

The raster layers are converted to polygons in order to do all the computations using only vector processing. Each cell of the raster is polygonized, and if an area of interest is provided, the polygons are clipped to it. This removes any unneeded grid cells. The resulting **HSgrid** object is a standard 'sf' object with information about each raster cell. The runoff timeseries (a [tsibble](https://github.com/tidyverts/tsibble)) can be found in a named list column *runoff_ts*. The elements are named by the ID, and can be accessed with the '$' notation.

We use **raster_to_HSgrid** to convert a raster timeseries to a HSgrid. Below, runoff is a *RasterBrick*, date is the date of the first layer in runoff, aoi is the area of interest.  
```{r message=FALSE, warning=FALSE}
HSgrid <- raster_to_HSgrid(runoff, 
                         date = ymd("1980-01-01"), 
                         timestep = "month", 
                         aoi = st_union(basins),
                         names = "LORA")
names(HSgrid)
plot(HSgrid[,"area_m2"])
HSgrid$runoff_ts$`1`
```

HSgrid can also be created from an 'sf' (polygon) object and a runoff timeseries using function [create_HSgrid](https://mkkallio.github.io/hydrostreamer/reference/create_HSweights.html).

## 2. Compute weights for downscaling

Once the raster has been converted to *HSgrid*, we can compute the weights. *compute_HSweights()* provides a possibility to weight the segments by their respective catchment area, or by using river segment properties. In this tutorial we will use the provided catchment areas, and the river network itself to compute the weights.

Below we provide five arguments to the segment weighting: *HSgrid* - which we just created in the previous step, *river* - the river network, *"length"* for weights - we'll use the river segment length as the basis of weighting, union of *basins* for area of interest - because we want only river segments within the area of interest. Finally, we tell the function that the river contains unique ID's in the column called *"SEGMENT_ID"*.

For catchment-based weighting, the only difference is that the weights argument gets value *"area"*, and we provide the *basins* object to the argument with the same name.
```{r message=FALSE, warning=FALSE}
river_weights <- compute_HSweights(HSgrid, 
                                   river, 
                                   weights = "length",
                                   aoi = st_union(basins),
                                   riverID = "SEGMENT_ID")

catchment_weights <- compute_HSweights(HSgrid, 
                                       river, 
                                       weights = "area",
                                       basins = basins,
                                       aoi = st_union(basins),
                                       riverID = "SEGMENT_ID")
```

To illustrate the differences between these two approaches, we can plot the **weights** element in the resulting *HSweights* object.

```{r}
par(mfrow=c(1,2))
plot(river_weights$weights[,"gridID"], reset=FALSE)
plot(st_union(basins), add=TRUE)
plot(catchment_weights$weights[,"gridID"])
```

The river segments and catchments are split at the borders of the runoff polygons. Weights are then divided among the river segments within each individual runoff polygon. The weights inside each runoff polygon always add to 1.

```{r}
par(mfrow=c(1,2))
plot(river_weights$weights[,"weights"], reset=FALSE)
plot(st_union(basins), add=TRUE)
plot(catchment_weights$weights[,"weights"])

river_weights$weights %>%
    group_by(gridID) %>%
    summarise(sum_of_weights = sum(weights))
```

The other elements of *HSweights* are the (routed) river network and the *HSgrid* object provided in the input to thhe constructing function. 

## 3. downscale runoff

Once a *HSweights* object has been created, we can apply the actual downscaling. The function condences the three elements of *HSweights* into an 'sf' object containing the river routing information, downscaled runoff timeseries (the unit is converted from a field, mm/s, to volume, m^3/s).

```{r message=FALSE, warning=FALSE}
catchment_downscaled <- downscale_runoff(catchment_weights)
( river_downscaled <- downscale_runoff(river_weights) )


```

The downscaled runoff in *runoff_ts* is a named list column where each element is a *tsibble*. Like with HSgrid, each element can be accessed by its name - the ID of the river segment.

```{r}
( runoff200 <- catchment_downscaled$runoff_ts$`200` )
```

We can compare the downscaled results from the two approaches.

```{r}
plotdata <- left_join(runoff200, 
                      river_downscaled$runoff_ts$`200`,
                      by="Date")
colnames(plotdata) <- c("Date", "Catchment", "River")
plotdata <- gather(plotdata, 
                   Approach,
                   Runoff,
                   -Date)

ggplot(plotdata, aes(Date, Runoff, color=Approach)) + 
    geom_line() +
    labs(title = "Downscaled runoff at river segment 200",
         y = 'Runoff in m3/s',
         x = NULL)
```


## 4. Apply river routing

While the downscaled runoff at each river segment is already useful for many applications, often knowing river discharge is also desirable. *hydrostreamer* provides three simple river routing algorithms for this purpose: **instantaneous** routing, useful for e.g. estimating runoff in the entire upstream catchment of each river segment, **simple** lag based routing, and **Muskingum** routing scheme.    

Each routing method is accessible through the function *accumulate_runoff()*. Since the catchment provided here is small, and the timestep in runoff is one month, we'll just use instantaneous routing here. There would be negligible difference between lag routing and instantaneous routing in this case.

```{r message=FALSE, warning=FALSE}
catchment_discharge <- accumulate_runoff(catchment_downscaled,
                                         method = "simple")
river_discharge <- accumulate_runoff(river_downscaled,
                                     method = "simple")
```

We can also plot the estimated discharge from the two approaches at segment 200 by accessing the new list column *discharge_ts*. 

```{r}
plotdata2 <- left_join(catchment_discharge$discharge_ts$`200`, 
                      river_discharge$discharge_ts$`200`,
                      by="Date")
colnames(plotdata2) <- c("Date", "Catchment", "River")
plotdata2 <- gather(plotdata2, 
                   Approach,
                   Discharge,
                   -Date)

ggplot(plotdata2, aes(Date, Discharge, color=Approach)) + 
    geom_line() +
    labs(title = "Downscaled discharge at river segment 200",
         y = 'Q m3/s',
         x = NULL)
```

As seen from the plots, the two approaches can lead into very different estimations in small streams - but as the catchment size increases, their predictions come closer and closer to one another.

We recommend that segment specific catchments be used instead of river lines for weighting whenever possible, because it creates more realistic division of runoff into the river segments.

## Exporting

*hydrostreamer* also contains a function to export the results as a GeoPackage, since it is much easier to explore the data in a GIS software with GUI. 
```{r message=FALSE, warning=FALSE, eval=FALSE}
# HSwrite(catchment_discharge, "downscaled_streamflow.gpkg")
# HSwrite(catchment_discharge, "downscaled_streamflow", what = "discharge")
```


## References

T. Tadono, H. Ishida, F. Oda, S. Naito, K. Minakawa, H. Iwamoto : Precise Global DEM Generation By ALOS PRISM, ISPRS Annals of the Photogrammetry, Remote Sensing and Spatial Information Sciences, Vol.II-4, pp.71-76, 2014. 

Hobeichi, S., Abramowitz, G., Evans, J., and Beck, H. E.: Linear Optimal Runoff Aggregate (LORA): a global gridded synthesis runoff product, Hydrol. Earth Syst. Sci., 23, 851-870, https://doi.org/10.5194/hess-23-851-2019, 2019.
